Accelerating adaptive ultrasound algorithms using graphics processing units \vspace{0.5cm}

### 2014-02-22 ThesisJP/Background/background.tex
Some background writing.

\section{General-purpose computing on graphics processing units}\label{sec:gpgpu}
In a recent essay, Herb Sutter, a leading authority on software development, reviews the current state of the software and computer industry \cite{HerbSutter}. The essay bears the name "Welcome to the Jungle" and is a sequel to the essay "The Free Lunch is Over" from 2004 \cite{HerbSuttera}. The two essays spins around the fact that chip manufactures hit a frequency-wall in the beginning of the 21st century. Until then, all mainstream computer programs where typically running in a single thread on a single core, and a programmer could expect the software to annually gain performance without touching the code. This was the "free-lunch"-era as depicted in Fig.\,\ref{fig:jungle}. The increase in processing power came mostly from an ever-growing clock frequency. However, power consumption and heat generation where also growing at the same rate, until the level of cooling required finally became to much in 2004.  From that point, chip manufactures had to pursue a different direction  than just increasing the clock frequency. The immediate solution was to embed several cores in one CPU chip. This insured continuation of Moore's law, since the number of transistors per chip could continue to grow. For programmers, multi-core CPUs meant that computer programs now had to be multi-threaded in order to continue to gain performance. It was a concurrency revolution, as Herb Sutter named it, and  the "free lunch" was over.

@@

For the first adopters of GPU programming, or General-purpose programming on graphics processing units (GPGPU) as it was named, it soon became evident that offloading all types of computations to the GPU was not always a good idea. In most cases it is important to balance the load equally between the CPU and GPU to obtain their combined computationally power. Balancing typically means that highly parallel tasks should be offloaded to the GPU and serialized work should stay on the CPU. This paradigm of utilizing specialized cores for solving specific problems is today known as heterogeneous computing. The specialized core can be a GPU or e.g. a field-programmable gate array (\nom{FPGA}{Field-programmable gate array}). A recent trend is that specialized accelerator cores are embedded onto the CPU. This heterogeneous processors is often referred to as an accelerated processing unit (\nom{APU}{Accelerated processing unit}). The last three generations of CPUs from Intel have all had on-chip integrated graphics. Even if an on-chip GPU is less powerful than a high-performance s discrete GPU, valuable time is saved by not having to send data across the PCI Express bus, which connects the different parts of a computer to the motherboard. This is the reason why the first rules of GPGPU (with discrete GPUs) is to minimize CPU-to-GPU transfers.

@@

\subsection{Comparing CPU and GPU performance}\label{sec:cpu_vs_gpu}
The CPU is design to be the main control unit of a computer. It is therefore good at general processing. The GPU, on the other hand, has a history of processing graphics which involves heavy vector arithmetic. Even though they started out as very different processors, modern CPUs are including more and more GPU-like features, and GPUs are adding CPU features. In this section we will see what these similarities are and why there is still a big difference in peak performance between a CPU and GPU.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{img/CPU-GPU.png}
\caption{Overview of CPU and GPU architecture. (make my own)}
\label{fig:cpu_gpu}
\end{figure}

In Section \ref{sec:gpgpu} we introduced a special type of instructions, namely SIMD instructions, which are designed to execute a single instruction across multiple data element in one clock cycle. At the end of the 20th century, Intel introduced a SIMD instruction set for their Pentium III processor called the Streaming SIMD Extensions (\nom{SSE}{Streaming SIMD extensions (instruction)}). This extension has seen several updates since. The first SEE instructions set had 128 bit long registers which meant that four 32 bit floating point values could be processed per clock cycle\footnote{For simplicity we will assume that the whole register is processed in one clock cycle even if the actual implementation might distributes one instruction for the register across multiple cycles.}. Updates of the SSE instructions sets have added new instructions (SSE2, SSE3, and SSE4) and longer registers (\nom{AVX}{Advanced vector extensions (instruction)}, AVX2, and AVX-512). AVX and AVX2 adds 256 bit long registers and fused multiply-accumulate instructions (\nom{FMA}{Fused multiply-accumulate (instruction)})  which means that sixteen 32 bit floating point values can be processed per clock cycle per core. A new extension, AVX-512, extends the registers to 512 bit and is found in the Intel accelerator board  Xeon Phi and might show up in the next generation of CPUs. The latest CPU architecture from Intel, the Haswell architecture, has two FMA ports and is therefore capable of processing 32, 32 bit, floating point values per clock cycle per core (Fig.\,\ref{fig:cpu_gpu}). For an eight core\footnote{An 8-core Haswell CPU is scheduled to launch during 2014.} CPU with a clock frequency of 3.8 GHz this sums to 
\begin{equation}
32*8*3.8 = 972.8\,\text{GFLOPS},
\end{equation}
hence a theoretical peak single precision performance of close to one tera single precision floating point operations per second (TFLOPS). This is the same number as visualized in Fig\,\ref{fig:cpu_vs_gpu} for the Haswell architecture. In addition, the most powerful integrated GPU found on Haswell CPUs have 40 execution units capable of processing sixteen, 32 bit, floating points at 1.1 GHz, 
\begin{equation}
16*40*1.1 = 704\,\text{GFLOPS}.
\end{equation}

Modern GPUs can also be interpreted as a unit consisting of multiple SIMD processors (Fig.\,\ref{fig:cpu_gpu}). The latest GPU architecture by Nvidia have up to 15 SIMD units capable of processing 192, 32 bit, floats per clock cycle at 745 MHz. This GPU is also capable of performing FMA instructions, doubling the number of theoretical FLOPS. The theoretical peak single precision FLOPS for this architecture is by that found to be 
\begin{equation}
15*192*2*745 = 4291\,\text{GFLOPS}.
\end{equation} 
Each SIMD core also comprise 16 special function units (\nom{SFU}{Special function unit (cos, exp, ect.)}) which add close to 1 TFLOPS in additional power. We see how the wide SIMD register found on GPUs results in five times the performance of a high-end CPU (without the integrated GPU).

Note that the derived numbers are strictly theoretical and that actual throughput typically is algorithm dependent. In this section we have focused on CPUs from Intel and GPUs from Nvidia, however, GPUs and CPUs from AMD have similar specifications.

@@

\subsection{Programming a GPU}
Early gpgpu programming required expert knowledge about graphics and the remaining fixed funtions in the pipeline. New programming languages where needed. OpenCL and CUDA. C++ AMP.

>>>>

\section{General-purpose computing on graphics processing units}\label{sec:gpgpu}
In a recent essay, Herb Sutter, a leading authority on software development, reviews the current state of the software and computer industry \cite{HerbSutter}. The essay bears the name "Welcome to the Jungle" and is a sequel to the essay "The Free Lunch is Over" from 2004 \cite{HerbSuttera}. The two essays spins around the fact that CPU manufactures hit a frequency-wall in the beginning of the 21st century. Until then, all mainstream computer programs where typically running in a single thread on a single CPU \textit{processing core}, and a programmer could expect the software to annually gain performance without touching the code. This was the "free-lunch"-era as depicted in Fig.\,\ref{fig:jungle}. The increase in processing power came mostly from an ever-growing clock frequency. However, power consumption and heat generation where also growing at the same rate, until the level of cooling required finally became to much in 2004.  From that point, chip manufactures had to pursue a different direction  than just increasing the clock frequency. The immediate solution was to embed several cores in one CPU chip. This insured continuation of Moore's law, since the number of transistors per chip could continue to grow. For programmers, multi-core CPUs meant that computer programs now had to be multi-threaded in order to annually gain performance. It was a concurrency revolution, as Herb Sutter named it, and  the "free lunch" was over.

@@

For the first adopters of GPU programming, or General-purpose programming on graphics processing units (GPGPU) as it was named, it soon became evident that offloading all types of computations to the GPU was not always a good idea. In most cases it is important to balance the load equally between the CPU and GPU to obtain their combined computationally power. Balancing typically means that highly parallel and computationally intense tasks should be offloaded to the GPU and serialized and memory intensive work should stay on the CPU. This paradigm of utilizing specialized cores for solving specific problems is today known as heterogeneous computing. The specialized core can be a GPU or e.g. a field-programmable gate array (\nom{FPGA}{Field-programmable gate array}). A recent trend is that specialized accelerator cores are embedded onto the CPU. This heterogeneous processors is often referred to as an accelerated processing unit (\nom{APU}{Accelerated processing unit}). The last three generations of CPUs from Intel have all had on-chip integrated graphics. Even if an on-chip GPU is less powerful than a high-performance s discrete GPU, valuable time is saved by not having to send data across the PCI Express bus, which connects the different parts of a computer to the motherboard. This is the reason why one of the first rules of GPGPU (with discrete GPUs) is to minimize CPU-to-GPU memory transfers.

@@

\subsection{Comparing CPU and GPU performance}\label{sec:cpu_vs_gpu}
The CPU is design to be the main control unit of a computer. It is therefore good at general processing. The GPU, on the other hand, has a history of processing graphics which involves heavy vector arithmetic. Even though they started out as very different processors, modern CPUs are including more and more GPU-like features, and GPUs are adding CPU features. In this section we will see what these similarities are and why there is still a big difference in peak performance between a CPU and a GPU.

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{img/CPU-GPU.pdf}
\caption{Schematic overview of CPU (Intel Haswell) and GPU (Nvidia Kepler) architecture. Note how more space is dedicated for compute cores on a GPU, and how more space is dedicated to control and caching on a CPU.}
\label{fig:cpu_gpu}
\end{figure}

In Section \ref{sec:gpgpu} we introduced a special type of instructions, namely SIMD instructions, which are designed to execute a single instruction across multiple data element in one clock cycle. At the end of the 20th century, Intel introduced a SIMD instruction set for their Pentium III processor called the Streaming SIMD Extensions (\nom{SSE}{Streaming SIMD extensions (instruction)}). This extension has seen several updates since. The first SEE instructions set had 128 bit long registers which meant that four 32 bit floating point values could be processed per clock cycle\footnote{For simplicity we will assume that the whole register is processed in one clock cycle even if the actual implementation might distributes one instruction for the register across multiple cycles.}. Updates of the SSE instruction set have added new instructions (SSE2, SSE3, and SSE4) and longer registers (\nom{AVX}{Advanced vector extensions (instruction)}, AVX2, and AVX-512). AVX and AVX2 adds 256 bit long registers and fused multiply-accumulate instructions (\nom{FMA}{Fused multiply-accumulate (instruction)})  which means that sixteen 32 bit floating point values can be processed per clock cycle per core. A new extension, AVX-512, extends the registers to 512 bit and is found in the Intel accelerator board  Xeon Phi and might show up in the next generation of CPUs. The latest CPU architecture from Intel, the Haswell architecture, has two processing ports supporting FMA-AVX2 instructions, and is therefore capable of processing 32, 32 bit, floating point values per clock cycle per core (Fig.\,\ref{fig:cpu_gpu}). For an eight core\footnote{An 8-core Haswell CPU is scheduled to launch during 2014.} CPU with a clock frequency of 3.8 GHz this sums to 
\begin{equation}
32*8*3.8 = 972.8\,\text{GFLOPS},
\end{equation}
hence a theoretical peak performance of close to one tera single precision floating point operations per second (TFLOPS). This is the same number as visualized in Fig\,\ref{fig:cpu_vs_gpu} for the Haswell architecture. In addition, the most powerful integrated GPU found on Haswell CPUs have 40 execution units capable of processing sixteen, 32 bit, floating points at 1.1 GHz. This adds
\begin{equation}
16*40*1.1 = 704\,\text{GFLOPS}
\end{equation}
in extra processing power to the CPU chip.

Modern GPUs can also be interpreted as a unit consisting of multiple SIMD processors (Fig.\,\ref{fig:cpu_gpu}). The latest GPU architecture by Nvidia have up to 15 SIMD units capable of processing 192, 32 bit, floats per clock cycle at 745 MHz. These GPUs are also capable of performing FMA instructions, effectively doubling the number of theoretical FLOPS. The theoretical peak single precision FLOPS for this architecture is by that found to be 
\begin{equation}
15*192*2*745 = 4291\,\text{GFLOPS}.
\end{equation} 
Each SIMD core also comprise 16 special function units (\nom{SFU}{Special function unit (sin, exp, sqrt, ect.)}) which add close to 1 TFLOPS in additional computational power. This adds up to five TFLOPS in computationally power as shown in Fig.\,\ref{fig:cpu_vs_gpu} for the GeForce GTX 780 TI GPU. From Fig.\,\ref{fig:cpu_gpu} we see how the high number of SIMD cores found on GPUs and their larger width results in five times the computational performance of a high-end CPU (without the integrated GPU).

Note that the derived numbers are strictly theoretical and that actual throughput is highly algorithm dependent. Nevertheless, in theory, the benefit of GPU computing over CPU computing is currently a factor of ten\footnote{A factor of five when a eight core Haswell CPU is launched later this year.} in raw computationally power. A claimed speed-up larger than this is typically caused by comparing a non-optimized CPU implementation with an optimized GPU implementation or by comparing uneven hardware \cite{Lee2010, Kothapalli2013}. Also note that in this section we have focused on CPUs from Intel and GPUs from Nvidia, however, CPUs and GPUs from AMD have similar specifications.

@@

\subsection{Programming a GPU}
Early GPGPU programming required expert knowledge about computer graphics and the remaining fixed functions in the pipeline. When GPGPU gained interest it quickly became evident that new programing languages where needed.  One of the first GPGPU oriented languages for the GPU was the Brook language by Buck \textit{et al.} \cite{Buck2004}. Buck later joined Nvidia where he led the work on \nom{CUDA}{Compute unified device architecture (GPGPU language)}, a GPGPU language by Nvidia for Nvidia GPUs only. The first version of CUDA was released in 2007. The following year, Apple launched their own GPGPU language called OpenCL, a standard which are administrated by the Krohnos Group.

New programming languages where needed. OpenCL and CUDA. C++ AMP.
